---
title: 'Common Distributions in R'
date: "Dr. K. Banner STAT 422 Notes (Fall 2020)"
output:
  pdf_document:
    includes:
      in_header: ../../notes/header-Banner.tex
  html_document:
    df_print: paged
fontsize: 12 pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, height = 5, width = 5, out.width="0.5\\linewidth", fig.align="center", collapse = TRUE)
```
<!-- See the .Rmd file in the course [Github repository](https://github.com/staceyhancock/stat501) for code to generate all figures. -->

#### Using \textsf{R} to further investigate the Binomial$(n,p)$ distribution

Make your best guess at the parameters (i.e., $n$ and $p$) of the following Binomial$(n,p)$ distributions. 

\vspace{.25in}

```{r, out.width = "90%", fig.width=8, fig.height=3, echo = FALSE}
par(mfrow = c(1,3))
x_obs <- 0:10
x1 <- dbinom(x = x_obs, prob = 0.2, size = 10)
x2 <- dbinom(x = x_obs, prob = 0.5, size = 10)
x3 <- dbinom(x = x_obs, prob = 0.9, size = 10)
x1_pmf <- rbind(x_obs, x1)
rownames(x1_pmf) <- c("x_1", "Pr(X_1 = x_1)")
x2_pmf <- rbind(x_obs, x2)
rownames(x2_pmf) <- c("x_2", "Pr(X_2 = x_2)")
x3_pmf <- rbind(x_obs, x3)
rownames(x2_pmf) <- c("x_3", "Pr(X_3 = x_3)")
plot(x = x_obs, y = x1, type = "h", xlab = expression(x[1]), 
     ylab = expression(f[X[1]](x[1])), lwd = 2)
points(x = x_obs, y = x1, pch = 20)
plot(x = x_obs, y = x2, type = "h", xlab = expression(x[2]), 
     ylab = expression(f[X[2]](x[2])), lwd = 2)
points(x = x_obs, y = x2, pch = 20)
plot(x = x_obs, y = x3, type = "h", xlab = expression(x[3]), 
     ylab = expression(f[X[3]](x[3])), lwd = 2)
points(x = x_obs, y = x3, pch = 20)
```

Use \textsf{R} to find values for Binomial probabilities. 

- To find $P(Y=y)$ use `dbinom(x = y, prob = p, size = n)` 
    ```{r}
# Example. Let Y ~ Binomial(n = 20, p = 0.2), find P(Y = 4)
dbinom(x = 4, size = 20, prob = 0.2)
    ```
- To find $P(Y\leq y)$ use `pbinom(q = y, prob = p, size = n)`
    ```{r}
# Example. Let Y ~ Binomial(n = 20, p = 0.2), find P(Y <= 4)
pbinom(q = 4, size = 20, prob = 0.2)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qbinom(p = c, size = n, prob = p)`
    ```{r}
# Example. Let Y ~ Binomial(n = 20, p = 0.2), find y such that P(Y <= y) = 0.5
qbinom(p = 0.5, size = 20, prob = 0.2)
    ```

- To simulate $m$ random draws from a Binomial($n$, $p$) distribution use     
`rbinom(n = m, size = n, prob = p)`
    ```{r}
# Example. Let Y ~ Binomial(n = 20, p = 0.2), generate a RS of size 5 
rbinom(n = 5, size = 20, prob = 0.2)
    ```

#### Using \textsf{R} to further investigate Geom$(p)$ RVs

Make your best guess at the parameter value (i.e., $p$).  

\vspace{.25in}

```{r, out.width = "100%", fig.width=8, fig.height=4, echo = FALSE}
par(mfrow = c(1,3))
x_obs <- 0:50
x1 <- dgeom(x = x_obs, prob = 0.5)
x2 <- dgeom(x = x_obs, prob = 0.1)
x3 <- dgeom(x = x_obs, prob = 0.8)
x1_pmf <- rbind(x_obs, x1)
rownames(x1_pmf) <- c("x_1", "Pr(X_1 = x_1)")
x2_pmf <- rbind(x_obs, x2)
rownames(x2_pmf) <- c("x_2", "Pr(X_2 = x_2)")
x3_pmf <- rbind(x_obs, x3)
rownames(x2_pmf) <- c("x_3", "Pr(X_3 = x_3)")
plot(x = x_obs + 1, y = x1, type = "h", xlab = expression(x[1]), 
     ylab = expression(f[X[1]](x[1])), lwd = 2)
points(x = x_obs + 1, y = x1, pch = 20)
plot(x = x_obs + 1, y = x2, type = "h", xlab = expression(x[2]), 
     ylab = expression(f[X[2]](x[2])), lwd = 2)
points(x = x_obs + 1, y = x2, pch = 20)
plot(x = x_obs + 1, y = x3, type = "h", xlab = expression(x[3]), 
     ylab = expression(f[X[3]](x[3])), lwd = 2)
points(x = x_obs + 1, y = x3, pch = 20)
```

What parameterization does \textsf{R} use for the geometric RV? 

\includegraphics[width=\textwidth]{img/Rgeom-param}


To use \textsf{R} to compute quantities of interst from the same Geometric($p$) distribution provided in the text, do the following:

- To find $P(Y=y)$ use `dgeom(x = y - 1, prob = p)` 
    ```{r}
# Example. Let Y ~Geom(p = 0.2), find P(Y = 4)
dgeom(x = 4 - 1, prob = 0.2)
    ```
- To find $P(Y\leq y)$ use `pgeom(q = y - 1, prob = p)`
    ```{r}
# Example. Let Y ~Geom(p = 0.2), find P(Y <= 4)
pgeom(q = 4 - 1, prob = 0.2)
    ```
\newpage
- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qgeom(p = c, prob = p) + 1`
    ```{r}
# Example. Let Y ~Geom(p = 0.2), find y such that P(Y <= y) = 0.5
qgeom(p = 0.5, prob = 0.2) + 1
    ```

- To simulate $m$ random draws from a Geom($p$) distribution use     
`rgeom(n = m, size = n) + 1`
    ```{r}
# Example. Let Y ~ Geom(p = 0.2), generate a RS of size 5 
rgeom(n = 5, prob = 0.2) + 1 
    ```
 
#### Using \textsf{R} to further investigate NegBinom$(r, p)$ RVs

Make your best guess at the parameter values (i.e., $r = ?$, $p$ = ?) for the following plots.   

\vspace{.25in}

```{r, out.width = "100%", fig.width=9, fig.height=4, echo = FALSE}
par(mfrow = c(1,3))
x_obs <- 0:100
x_r1 <- x_obs + 7
x_r <- x_obs + 4
x_r3 <- x_obs + 9
x1 <- dnbinom(x = x_obs[1:20], size = 7, prob = 0.5)
x2 <- dnbinom(x = x_obs, size = 4, prob = 0.1)
x3 <- dnbinom(x = x_obs[1:20], size = 9, prob = 0.8)
x1_pmf <- rbind(x_obs[1:20], x1)
rownames(x1_pmf) <- c("x_1", "Pr(X_1 = x_1)")
x2_pmf <- rbind(x_obs, x2)
rownames(x2_pmf) <- c("x_2", "Pr(X_2 = x_2)")
x3_pmf <- rbind(x_obs[1:20], x3)
rownames(x2_pmf) <- c("x_3", "Pr(X_3 = x_3)")
plot(x = x_r1[1:20], y = x1, type = "h", xlab = expression(x[1]), 
     ylab = expression(f[X[1]](x[1])), lwd = 2)
points(x = x_r1[1:20], y = x1, pch = 20)
plot(x = x_r, y = x2, type = "h", xlab = expression(x[2]), 
     ylab = expression(f[X[2]](x[2])), lwd = 2, xlim = c(4, 104))
abline(v = 4, lty = 2)
text(x = 4, y = 0.022, labels = bquote(x[2]~"= 4"), pos = 4)
points(x = x_r, y = x2, pch = 20)
plot(x = x_r3[1:20], y = x3, type = "h", xlab = expression(x[3]), 
     ylab = expression(f[X[3]](x[3])), lwd = 2)
points(x = x_r3[1:20], y = x3, pch = 20)
```

Just like `_geom`, `_nbinom` is parameterized differently in \textsf{R}! 

\includegraphics[width=\textwidth]{img/nbinom-param}

To use \textsf{R} to compute quantities of interst from the same parameterization of the NegativeBinomial($r$, $p$) distribution that is provided in the text, do the following:

- To find $P(Y=y)$ use `dnbinom(x = y - r, size = r, prob = p)` 
    ```{r}
# Example. Y ~ NegBinom(r = 4, p = 0.2), find P(Y = 8)
dnbinom(x = 8 - 4, size = 4, prob = 0.2)
    ```
- To find $P(Y\leq y)$ use `pgeom(q = y, size = r, prob = p)`
    ```{r}
# Example. Let Y ~ NegBinom(r = 4, p = 0.2), find P(Y <= 8)
pnbinom(q = 8 - 4, size = 4, prob = 0.2)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qnbinom(p = c, size = r, prob = p) + r`
    ```{r}
# Example. Let Y ~ NegBinom(r = 4, p = 0.2), find y such that P(Y <= y) = 0.5
qnbinom(p = 0.5, size = 4, prob = 0.2) + 4 
    ```

- To simulate $m$ random draws from a NegBinom($r$, $p$) distribution use    
`rnbinom(n = m, size = r, prob = p) + r`
    ```{r}
# Example. Let Y ~ NegBinom(r = 4, p = 0.2), generate a RS of size 5 
rnbinom(n = 5, size = 4, prob = 0.2) + 4 
    ```


#### Using \textsf{R} to further investigate Hypergeometric$(N,r,n)$ RVs

Make your best guess at the parameter values (i.e., $N = 30$, $r = 18$, $n$ = ?) for the following plots.   

\vspace{.25in}

```{r, out.width = "100%", fig.width=9, fig.height=4, echo = FALSE}
par(mfrow = c(1,3))
x_obs <- 0:12
x1 <- dhyper(x = x_obs, m = 12, n = 18, k = 10)
x2 <- dhyper(x = x_obs, m = 12, n = 18, k = 15)
x3 <- dhyper(x = x_obs, m = 12, n = 18, k = 5)
x1_pmf <- rbind(x_obs, x1)
rownames(x1_pmf) <- c("x_1", "Pr(X_1 = x_1)")
x2_pmf <- rbind(x_obs, x2)
rownames(x2_pmf) <- c("x_2", "Pr(X_2 = x_2)")
x3_pmf <- rbind(x_obs, x3)
rownames(x2_pmf) <- c("x_3", "Pr(X_3 = x_3)")
plot(x = x_obs[1:11], y = x1[which(x1 != 0)], type = "h", xlab = expression(x[1]), 
     ylab = expression(f[X[1]](x[1])), lwd = 2)
points(x = x_obs, y = x1, pch = 20)
plot(x = x_obs, y = x2, type = "h", xlab = expression(x[2]), 
     ylab = expression(f[X[2]](x[2])), lwd = 2)
points(x = x_obs, y = x2, pch = 20)
plot(x = x_obs[1:6], y = x3[which(x3 != 0)], type = "h", xlab = expression(x[3]), 
     ylab = expression(f[X[3]](x[3])), lwd = 2)
points(x = x_obs, y = x3, pch = 20)
```

\newpage

The help file for `_hyper` is shown below, what does each argument provide for the conventional notation $n$ = sample size, $N$ = total in population, $M$ = total number of successes. 


\includegraphics[width=\textwidth]{img/r-hypergeometric}


- To find $P(Y=y)$ use `dhyper(x = y, m = M, n = N-M, k = n)` 
    ```{r}
# Example. Y ~ Hyper(N = 10, r = 4, n = 5), find P(Y = 3)
dhyper(x = 3, m = 4, n = 6, k = 5)
    ```
- To find $P(Y\leq y)$ use `phyper(q = y, m = M, n = N-M, k = n)`
    ```{r}
# Example. Y ~ Hyper(N = 10, r = 4, n = 5), find P(Y <= 3)
phyper(q = 3, m = 4, n = 6, k = 5)
sum(dhyper(x = 0:3, m = 4, n = 6, k = 5))
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qhyper(p = c, m = M, n = N-M, k = n)`
    ```{r}
# Example. Let Y ~ Hyper(N = 10, r = 4, n = 5), find y such that P(Y <= y) = 0.5
qhyper(p = 0.5, m = 4, n = 6, k = 5)
phyper(q = 2, m = 4, n = 6, k = 5)
phyper(q = 1, m = 4, n = 6, k = 5)
    ```

- To simulate $m$ random draws from a Hypergeometric($N$, $M$, $n$) distribution use    
`rhyper(n = m, m = M, n = N-M, k = n) `
    ```{r}
# Example. Let  Y ~ Hyper(N = 10, r = 4, n = 5), generate a RS of size 5 
rhyper(nn = 5, m = 4, n = 6, k = 5) 
    ```

\newpage

#### Using \textsf{R} to further investigate Poisson$(\lambda)$ RVs

Make your best guess at the parameter value, $\lambda$ for the following plots.   

\vspace{.25in}

```{r, out.width = "100%", fig.width=9, fig.height=4, echo = FALSE}
par(mfrow = c(1,3))
x_obs <- 0:100
x1 <- dpois(x = x_obs, lambda = 5)
x2 <- dpois(x = x_obs, lambda = 12)
x3 <- dpois(x = x_obs, lambda = 75)
x1_pmf <- rbind(x_obs, x1)
rownames(x1_pmf) <- c("x_1", "Pr(X_1 = x_1)")
x2_pmf <- rbind(x_obs, x2)
rownames(x2_pmf) <- c("x_2", "Pr(X_2 = x_2)")
x3_pmf <- rbind(x_obs, x3)
rownames(x2_pmf) <- c("x_3", "Pr(X_3 = x_3)")
plot(x = x_obs[1:20], y = x1[1:20], type = "h", xlab = expression(x[1]), 
     ylab = expression(f[X[1]](x[1])), lwd = 2)
points(x = x_obs, y = x1, pch = 20)
plot(x = x_obs[1:40], y = x2[1:40], type = "h", xlab = expression(x[2]), 
     ylab = expression(f[X[2]](x[2])), lwd = 2)
points(x = x_obs, y = x2, pch = 20)
plot(x = x_obs, y = x3, type = "h", xlab = expression(x[3]), 
     ylab = expression(f[X[3]](x[3])), lwd = 2)
points(x = x_obs, y = x3, pch = 20)
```

- To find $P(Y=y)$ use `dpois(x = y, lambda = lambda)` 
    ```{r}
# Example. Y ~ Poisson(lambda = 5), find P(Y = 3)
dpois(x = 3, lambda = 5)
    ```
- To find $P(Y\leq y)$ use `ppois(q = y, lambda = lambda)`
    ```{r}
# Example. Y ~ Poisson(lambda = 5), find P(Y <= 3)
ppois(q = 3, lambda = 5)
sum(dpois(x = 0:3, lambda = 5))
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qpois(p = c, lambda = lambda)`
    ```{r}
# Example. Let Y ~ Poisson(lambda = 5), find y such that P(Y <= y) = 0.5
qpois(p = 0.5, lambda = 5)
ppois(q = 5, lambda = 5)
ppois(q = 4,lambda = 5)
    ```

- To simulate $m$ random draws from a Poisson($\lambda$) distribution use    
`rpois(n = m, lambda = lambda) `
    ```{r}
# Example. Let  Y ~ Poisson(lambda = 5), generate a RS of size 5 
rpois(n = 5, lambda = 5) 
    ```

## Some common continuous distributions

#### Using \textsf{R} to further investigate the uniform distribution

Find the parameters of the following Uniform$(\theta_1, \theta_2)$ distributions. That is, find $\theta_1$ and $\theta_2$. Come up with an example of a RV that could be modeled by one of the distributions below.

\vspace{.25in}

```{r, out.width = "0.95\\linewidth", fig.width=12, fig.height=6, echo = FALSE}
par(mfrow = c(1,3))
curve(dunif(x, min = 0, max = 60), from = -10, to = 70, 
      ylab = expression(f[X](x)))
curve(dunif(x, min = 0, max = 1), from = -0.25, to = 1.25, 
      ylab = expression(f[X](x)))
curve(dunif(x, -10,10), from = -11, to = 11, 
      ylab = expression(f[X](x)))
```

Using \textsf{R} to find values for uniform probabilities is a bit overkill due to the shape of the density, but here's how you do it. Let $Y \sim Uniform(\theta_1,\theta_2)$

- To find the _density_ of the pdf (i.e., the height of the pdf) at a particular value for $y$, use `dunif(x = y, min = theta1, max = theta2)`. What is $P(Y = y)$?
    ```{r}
# Example. Let Y ~ Uniform(0,1), find density at y = 0.5 
dunif(x = 0.5, min = 0, max = 1)
    ```
- To find $P(Y\leq y)$ use `punif(q = y, min = theta1, max = theta2)`
    ```{r}
# Example.  Let Y ~ Uniform(0,1), find P(Y <= 0.5)
punif(q = 0.5, min = 0, max = 1)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qunif(p = c, min = theta1, max = theta2)`, or $\phi_c$ 
    ```{r}
# Example. Let Y ~ Uniform(0,1), y such that P(Y <= y) = 0.5
qunif(p = 0.5, min = 0, max = 1)
    ```

- To simulate $m$ random draws from a Uniform($\theta_1$, $\theta_2$) distribution use
`runif(n = m, min = theta1, max = theta2)`. This is the most useful function for the uniform distribution! 
    ```{r}
# Example. Let Y ~ Uniform(0,1), generate a RS of size 5
runif(n = 5, min = 0, max = 1)
    ```

#### The normal distribution 

```{r, out.width = "90%", fig.width=9, fig.height=4, echo = FALSE}
par(mfrow = c(1,3))
curve(dnorm(x, 0,1), from = -10, to = 10, 
      ylab = expression(f[X](x)))
curve(dnorm(x, 0,3), from = -10, to = 10, 
      ylab = expression(f[X](x)))
curve(dnorm(x, 0,0.5), from = -10, to = 10, 
      ylab = expression(f[X](x)))
```

Using \textsf{R} to find normal probabilities can save you from transforming to a standard normal and then using a probability table! 

- To find the _density_ of the pdf (i.e., the height of the pdf) at a particular value for $y$, use `dnorm(x = y,  mean = mu, sd = sigma)`. Recall, $P(Y = y)$ = 0 $\forall y \in \mathbb{R}$!
    ```{r}
# Example. Let Y ~ Normal(10,4), find density at y = 5
dnorm(x = 5, mean = 10, sd = 4)
    ```
- To find $P(Y\leq y)$ use `pnorm(q = y, mean = mu, sd = sigma)`
    ```{r}
# Example. Let Y ~ Normal(10,4), P(Y <=5)
pnorm(q = 5, mean = 10, sd = 4)
# convert to z-score 
z <- (5 - 10)/4
z
pnorm(q = z, mean = 0, sd = 1)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qnorm(p = c, mean = mu, sd = sigma)`, or $\phi_c$ 
    ```{r}
# Example. Let Y ~ Normal(10,4), y such that P(Y <= y) = 0.25
qnorm(p = 0.25, mean = 10, sd = 4)
    ```

- To simulate $m$ random draws from a Normal($\mu$, $\sigma$) distribution use
`rnorm(n = m, mean = mu, sd = sigma)`. This can be incredibly useful for simulation studies! 
    ```{r}
# Example. Let Y ~ Normal(10,4), generate a RS of size 5
rnorm(n = 5, mean = 10, sd = 4)
    ```

\newpage

#### The gamma family of distributions 

Gamma distributions are incredibly flexible in shape. 

```{r echo = FALSE, fig.width=9, fig.height=4, out.width = "0.8\\linewidth", fig.align="center"}
par(mfrow = c(1,2))

plot(c(1,1), xlim = c(0,20), ylim = c(0,0.8), type = "n", 
	  ylab = expression(f[Y](y)), xlab = "y", 
	  main = bquote("Gamma("~alpha~","~beta~" = 1 )"))

# different alphas fixed beta
curve(dgamma(x, shape = 0.5, rate = 1), add = T, lwd = 3, col = "#377eb8")
curve(dgamma(x, shape = 2, rate = 1), add = T, lwd = 3, col = "#4daf4a")
curve(dgamma(x, shape = 5, rate = 1), add = T, lwd = 3, col = "#984ea3")
curve(dgamma(x, shape = 10, rate = 1), add = T, lwd  = 3, col = "#ff7f00")

legend("topright", bty = "n", legend = expression(alpha == 0.5, 
												   alpha == 2,
												   alpha == 5, 
												   alpha == 10), 
		merge= TRUE, col = c('#377eb8', '#4daf4a',
							 '#984ea3','#ff7f00'), 
		lty = rep(1,4), lwd = rep(3,4))
		


plot(c(1,1), xlim = c(0,40), ylim = c(0,0.8), type = "n", 
	  ylab = expression(f[Y](y)), xlab = "y", 
	  main = bquote("Gamma("~alpha~" = 1,"~beta~")"))

# colors came from colorbrewer2.org
# different betas fixed alpha

curve(dgamma(x, shape = 2, rate = 1/0.5), add = T, lwd = 3, col = "#cab2d6")
curve(dgamma(x, shape = 2, rate = 1/2), add = T, lwd = 3, col = "#a65628")
curve(dgamma(x, shape = 5, rate = 1/5), add = T, lwd = 3, col = "#f781bf")

legend("topright", bty = "n", legend = expression(beta == 0.5, 
												   beta == 2,
												   beta == 5), 
		merge= TRUE, col = c('#cab2d6', '#a65628',
							 '#f781bf'), 
		lty = rep(1,4), lwd = rep(3,3))
```

#### Using \textsf{R}  to find gamma probabilities

The functions `_gamma()`, `_exp()` and `_chisq()` can be used with `d`, `p`, `q`, and `r` to find values of the _pdf_, _CDF_, quantiles, and generate RS from the distributions, respectively. Just like the geometric and negative binomial distributions, the parameterization in \textsf{R} is different than what is in the text. Let $Y \sim Gamma(\alpha, \beta)$. 

- To find the _density_ of the pdf (i.e., the height of the pdf) at a particular value for $y$, use `dgamma(x = y,  shape = alpha, rate = 1/beta)`. Recall, $P(Y = y)$ = 0 $\forall y \in \mathbb{R}$!
    ```{r, results = "hide"}
# Example. Let Y ~ Gamma(4,5), find density at y = 5
dgamma(x = 5, shape = 4, rate = 1/5)
# for Y ~ Exp(5)
dexp(x = 5, rate = 1/5)
# for Y ~ chisquared(nu = 5)
dchisq(x = 5, df = 5)
    ```
- To find $P(Y\leq y)$ use `pgamma(q = y, shape = alpha, rate = 1/beta)`
    ```{r, results = "hide"}
# Example. Let Y ~ Gamma(4,5), P(Y <=5)
pgamma(q = 5, shape = 4, rate = 1/5)
# for Y ~ Exp(5)
pexp(q = 5, rate = 1/5)
# for Y ~ chisquared(nu = 5)
pchisq(q = 5, df = 5)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qgamma(p = c, shape = alpha, rate = 1/beta)`, or $\phi_c$ 
    ```{r, results = "hide"}
# Example. Let Y ~ Gamma(4,5), y such that P(Y <= y) = 0.25
qgamma(p = 0.25, shape = 4, rate = 1/5)
# for Y ~ Exp(5)
qexp(p = 0.25, rate = 1/5)
# for Y ~ chisquared(nu = 5)
qchisq(p = 0.25, df = 5)
    ```

\newpage

- To simulate $m$ random draws from a Gamma distribution use
`rgamma(n = m, shape = alpha, rate = 1/beta)`.
    ```{r, results = "hide"}
# Example. Let Y ~ Gamma(4,5), generate a RS of size 5
rgamma(n = 5, shape = 4, rate = 1/5)
# for Y ~ Exp(5)
rexp(n = 5, rate = 1/5)
# for Y ~ chisquared(nu = 5)
rchisq(n = 5, df = 5)
    ```

```{r, out.width = "0.95\\linewidth", fig.width=12, fig.height=6, echo = FALSE}
par(mfrow = c(1,3))
curve(dgamma(x, shape = 4, rate = 1/5), from = 0, to = 60, 
      ylab = expression(f[X](x)), 
      main = bquote("Gamma("~ alpha~"= 4"~","~beta~"= 5"~")"))
curve(dexp(x, rate = 1/5), from = 0, to = 20,
      ylab = expression(f[X](x)),
      main = bquote("Exp("~beta~"= 5"~")"))
curve(dgamma(x, shape = 5/2, rate = 1/2), from = 0, to = 15, 
      ylab = expression(f[X](x)), 
      main = bquote("Gamma("~ alpha~"= 5/2"~","~beta~"= 2"~")"))
```

\newpage

#### The beta distribution in \textsf{R} 

```{r, out.width = "0.95\\linewidth", fig.width=12, fig.height=6, echo = FALSE}
par(mfrow = c(1,3))
curve(dbeta(x, shape1 = 2, shape2 = 10), from = 0, to = 1, 
      ylab = expression(f[X](x)), 
      main = bquote("Beta("~ alpha~"= 2"~","~beta~"= 10"~")"))
curve(dbeta(x, shape1 = 1/2, shape2 = 1/2), from = 0, to = 1,
      ylab = expression(f[X](x)),
      main = bquote("Beta("~ alpha~"= 1/2"~","~beta~"= 1/2"~")"))
curve(dbeta(x, shape1 = 1, shape2 = 1), from = 0, to = 1, 
      ylab = expression(f[X](x)), 
      main = bquote("Beta("~ alpha~"= 1/2"~","~beta~"= 1/2"~")"))
```

The functions `_beta` can be used with `d`, `p`, `q`, and `r` to find values of the _pdf_, _CDF_, quantiles, and generate RS from the beta distribution, respectively. Let $Y \sim Beta(\alpha, \beta)$. 

- To find the _density_ of the pdf (i.e., the height of the pdf) at a particular value for $y$, use `dbeta(x = y,  shape1 = alpha, shape2 = beta)`. Recall, $P(Y = y)$ = 0 $\forall y \in \mathbb{R}$!
    ```{r}
# Example. Let Y ~ Beta(2,10), find density at y = 0.5
dbeta(x = 0.5, shape1 = 2, shape2 = 10)
    ```
- To find $P(Y\leq y)$ use `pbeta(q = y,  shape1 = alpha, shape2 = beta)`
    ```{r}
# Example. Let Y ~ Beta(2,10), P(Y <=0.5)
pbeta(q = 0.5, shape1 = 2, shape2 = 10)
    ```

- To find the smallest $y$ such that $P(Y\leq y) \geq c$ use `qbeta(p = c,  shape1 = alpha, shape2 = beta)`, or $\phi_c$ 
    ```{r}
# Example. Let Y ~ Beta(2,10), y such that P(Y <= y) = 0.25
qbeta(p = 0.25,  shape1 = 2, shape2 = 10)
    ```

- To simulate $m$ random draws from a Gamma distribution use
`rbeta(n = m,  shape1 = alpha, shape2 = beta)`.
    ```{r}
# Example. Let Y ~ Beta(2,10), generate a RS of size 5
rbeta(n = 5,  shape1 = 2, shape2 = 10)
    ```
